Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-06 ms/manuscript.Rmd
Use lajolla palette

**\fig \@ref(fig:tree)**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Relative value of quantitative (sporophyte) traits in heat gradient: low values are lighter, high values are darker. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Trends for three leaf size traits (stipe length, frond length, and frond with) were quantitatively very similar, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see \tab \@ref(tab:traits-used). Major Cretaceous epiphytic radiations \sensu @Schuettpelz2009 labeled on tree in green: H, hymenophylloids; T, trichomanoids, V, vittarioids; A, asplenioids; E, elaphoglossoids; P, polygrammoids (trichomanoids diversified prior to the Cretaceous but are included as an important extant epiphytic clade).

@@

```{r phylosig-cont}
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in continuous traits", format = "latex", booktabs = TRUE)
```

@@

```{r phylosig-binary}
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  select(trait, `Number of presences` = num_present, `Number of absences` = num_absent,
         P_rnd = prob_random, P_bm = prob_brownian) %>%
column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in binary traits", format = "latex", booktabs = TRUE)
```

@@

```{r pics}
pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth", format = "latex", booktabs = TRUE) %>%
  footnote(general = "All trait contrasts were made by subtracting epiphytic values from terrestrial values; thus, a positive contrast indicates greater values for terrestrial clades.", general_title = "", threeparttable = TRUE)
```

@@

```{r corr-evo}
correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Pagel's (1994) test of correlated evolution between binary traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial). Lower log likelihood for the dependent model indicates correlation with growth habit.", general_title = "", threeparttable = TRUE)
```

>>>>

**\fig \@ref(fig:tree)**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see \tab \@ref(tab:traits-used). Major Cretaceous epiphytic radiations \sensu @Schuettpelz2009 labeled on tree in green: H, hymenophylloids; T, trichomanoids, V, vittarioids; A, asplenioids; E, elaphoglossoids; P, polygrammoids (trichomanoids diversified prior to the Cretaceous but are included as an important extant epiphytic clade). Units on timescale for tree in millions of years.

@@

```{r phylosig-cont}
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait) %>%
  rename(K = kval, `P(K)` = k_pval, `P(lambda)` = lambda_pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in continuous traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "K, Blomberg's \\\\textit{K} (Blomberg \\\\textit{et al.}, 2003); lambda, Pagel's lambda (Pagel, 1999). Values of lambda or \\\\textit{K} greater than 1 indicate traits evolving according to Brownian Motion (BM); values of lambda or K near 0 indicate random distribution of traits.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

@@

```{r phylosig-binary}
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  select(trait, `Number of presences` = num_present, `Number of absences` = num_absent,
         D, `P(rnd)` = prob_random, `P(bm)` = prob_brownian) %>%
column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in binary traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "D, Fritz and Purvis’ \\\\textit{D} (Fritz \\\\& Purvis, 2010); P(rnd), probability of random distribution of traits; P(bm), probability of traits evolving under Brownian motion (BM). Values of \\\\textit{D} 0 or less indicate evolution according to BM; values near 1 indicate random trait distribution.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

@@

```{r pics}
pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait) %>%
  rename(t = statistic, P = p.value, `Num. constrasts` = n_contrasts, `Num. pos. constrasts` = n_pos_con) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth", format = "latex", booktabs = TRUE) %>%
  footnote(general = "All trait contrasts were made by subtracting epiphytic values from terrestrial values; thus, a positive contrast indicates greater values for terrestrial clades.", general_title = "", threeparttable = TRUE)
```

@@

```{r corr-evo}
correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  rename(`LogL(independent)` = logL_indep, `LogL(dependent)` = logL_dep, `Likelihood ratio` = likelihood_ratio, P = pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Pagel's (1994) test of correlated evolution between binary traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "LogL, log-likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial growth). Traits with a significantly lower log likelihood for the dependent model are significantly correlated with growth habit.", general_title = "", threeparttable = TRUE)
```
